Scoring Patients

### 2017-05-16 Patient Report.Rmd
fixing loads of little mistakes that results in messing up the pipeline entirely.

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
library(xlsx)
library(dplyr)
library(pander)
library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(tibble)

@@

## Patient info file
patient.file = grep('All_patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

dataset = 'Training'
patient.info = read.patient.info(patient.file, set=dataset)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

@@

sum.patient.data = summarise.patient.info(patient.info)

@@

@@

for (patient.name in unique(patient.info$Patient)  )  {
  message(patient.name)
  patient.plot.dir = paste(plot.dir, patient.name, sep='/')

@@

  if (length(missing) > 0) 
    warning(paste("Missing", length(missing), "sample(s) in patient", patient.name))

  message( paste(length(which(segvals$n.probes < min.probes)), ' probes below the minimum probe count (',min.probes,')', sep=''))

  segvals = segvals[segvals$n.probes>=min.probes,]

@@

@@

  if(nrow(segvals)>0) {
      HC = hclust(dist(t(segvals[,-(1:5)])))
      gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
      #print(gg1)

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,-(1:5)]
      for(c in 1:nrow(normalised.segvals)) {
        normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))
      }
      patient.data[[patient.name]][['norm.seg.vals']] = normalised.segvals

      # Variable Regions
      sds = apply(segvals[,-(1:5)],1,sd)
      means = apply(segvals[,-(1:5)],1,mean)
      CofV = sds/means

@@

      gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')
     # print(gg3)
      #pt.plots[[patient.name]][['norm.hist']] = gg

      # par(mfrow=c(3, no.rows))
      # for(s in 1:no.samples) 
      #   hist(normalised.segvals[,s],col="blue",main=names(normalised.segvals)[s],xlab="normalised coverage")

      if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology, Samplename)
        }
        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]
      }
}

sum.patient.data = subset(sum.patient.data, Patient %in% names(patient.data))

## Check that all of the samples are there
patient.info = arrange(patient.info, Status, Patient, Endoscopy.Year,Samplename)
pt = 'AH0254'
info = subset(patient.info, Patient == 'AH0254')
lapply(names(patient.data), function(pt) {
  missing = which(!patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals))
  if (length(missing) > 0) {
    msg = paste(pt,": Missing samples - ", paste(patient.data[[pt]]$info$Samplename[missing], collapse=', ') , sep='')
    msg = paste(msg, "\nFrom batch: ", paste(unique(patient.data[[pt]]$info$Batch.Name[missing]), collapse=', '))
    warning(msg)
  }
})

@@

example.pt = subset(sum.patient.data, Status == 'P' & total.samples > 2)$Patient[1]
## Note the number of probes per patient/sample
```
\pagebreak

@@

## Per-sample normalization
#nsv = patient.data[['AD0258']]$norm.seg.vals
#nsv[] = lapply(patient.data[['AD0258']]$seg.vals[,-(1:5)], function(x) (x-mean(x))/sd(x))

#ggplot(melt(nsv), aes(x=value))  + 
#        facet_wrap(~variable,scales = "free_x") + 
#        geom_histogram( fill=I("blue"),col=I("lightblue"), binwidth=0.3) + labs(title='AD0258 normalized segmented value distribution', x='', y='')

@@

pander(sum.patient.data[,c('Patient','Status','years','total.samples','highest.path','n.segments', 'median.seg.len')], justify='left')

#seg.lens = lapply(patient.data, function(df) df$seg.vals[,c('end.pos')] - df$seg.vals[,c('start.pos')] )

#wilcox.test(log(unlist(seg.lens[subset(sum.patient.data, Status == 'P')$Patient])),
#log(unlist(seg.lens[subset(sum.patient.data, Status == 'NP')$Patient])))

#hist(log(unlist(seg.lens[subset(sum.patient.data, Status == 'P')$Patient])))
#hist(log(unlist(seg.lens[subset(sum.patient.data, Status == 'NP')$Patient])))

ggplot(sum.patient.data) + geom_boxplot( aes(y=median.seg.len, x=Status, group=Status, fill=Status) ) +
  labs(title="Median segment length by patient status")

@@

# Segmented Value Distributions
```{r echo=F, message=F, warning=F, fig.height=15, fig.width=10, eval=F}

lapply(patient.data, function(pt) {
  name = pt$info$Patient[1]
  status = pt$info$Status[1]

@@

>>>>

```{r setup, include=FALSE}
library(ggplot2)
library(GGally)
library(plyr)
#library(xlsx)
library(dplyr)
library(pander)
#library(Hmisc)
library(ggdendro)
library(reshape2)
library(GenomicRanges)
library(gridExtra)
library(tibble)

@@

## Patient info file
patient.file = grep('/All_patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

dataset = 'Training'
#dataset = 'Test'
patient.info = read.patient.info(patient.file, set=dataset)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

bad_batch = which(patient.info$Batch.Name %in% c('Batch_8', 'Batch_9'))
patient.info = patient.info[-(bad_batch),]

@@

sum.patient.data = summarise.patient.info(patient.info)
table(sum.patient.data$Status)

@@

lowprobes = c()

@@

for (patient.name in sum.patient.data$Patient  )  {
  message(patient.name)
  patient.plot.dir = paste(plot.dir, patient.name, sep='/')

@@

  if (length(missing) > 0) {
    warning(paste("Missing", length(missing), "sample(s) in patient", patient.name))
    next
  }

  probes = length(which(segvals$n.probes < min.probes))
  message( paste(probes, ' probes (', round(probes/nrow(segvals))*100, '%)',' below the minimum probe count (',min.probes,')', sep=''))
  if (probes/nrow(segvals) >= 0.75) lowprobes = c(lowprobes, patient.name)

  segvals = segvals[segvals$n.probes >= min.probes,]
  dim(segvals)

@@

  if (nrow(segvals) < 100) {
    warning(paste(patient.name, 'has fewer than 100 genomic reginos with segmented values:', nrow(segvals)))
    next
  }

@@

  sample.cols = grep('^D\\d+_D\\d+', colnames(segvals))
  if (nrow(segvals) > 0) {
      HC = hclust(dist(t(segvals[,sample.cols])))
      gg1 = ggdendrogram(HC, rotate=T) + labs(title = paste(patient.name,": fitted segment coverage"),x="")
      #print(gg1)

      # Normalize  (value-mean(value))/sd(value)
      normalised.segvals = segvals[,sample.cols]
      for(c in 1:nrow(normalised.segvals)) 
        normalised.segvals[c,] = (normalised.segvals[c,]-mean(unlist(normalised.segvals[c,])))/sd(unlist(normalised.segvals[c,]))

      patient.data[[patient.name]][['norm.seg.vals']] = normalised.segvals

      # Variable Regions
      sds = apply(segvals[,sample.cols],1,sd)
      means = apply(segvals[,sample.cols],1,mean)
      CofV = sds/means

@@

      gg3 = ggplot( melt(normalised.segvals), aes(x=value) ) + 
        facet_wrap(~variable,scales = "free_x") + 
        geom_histogram( fill=I("blue"),col=I("lightblue"),binwidth=0.3) + labs(title=paste(patient.name,'normalised segment coverage'), x='', y='')

      #if(nrow(segvals)>=3) {
        # order by date of endoscopy
        patient = subset(patient.info, Patient == patient.name)
        patient = arrange(patient, Endoscopy.Year, Pathology, Samplename)
       # }
        sample.normalised.segvals = normalised.segvals[, intersect(colnames(normalised.segvals), patient$Samplename)]
      }
}

print(paste("Missing patients:", paste(setdiff(sum.patient.data$Patient, names(patient.data)), collapse=', ')))

sum.patient.data = subset(sum.patient.data, Patient %in% names(patient.data))
table(sum.patient.data$Status)

## Check that all of the samples are there
patient.info = arrange(patient.info, Status, Patient, Endoscopy.Year,Samplename)
msgs = sapply(names(patient.data), function(pt) {
  missing = which(!patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals))
  if (length(missing) > 0) {
    msg = paste(pt,": Missing samples - ", paste(patient.data[[pt]]$info$Samplename[missing], collapse=', ') , sep='')
    msg = paste(msg, "\nFrom batch: ", paste(unique(patient.data[[pt]]$info$Batch.Name[missing]), collapse=', '))
    warning(msg)
  }
})
for (n in msgs) {
  if (!is.null(n))
    print(n)
}

@@

example.pt = subset(sum.patient.data, Status == 'P' & total.samples > 2)$Patient[5]
## Note the number of probes per patient/sample
```
\pagebreak

@@

@@

ggplot(sum.patient.data) + geom_boxplot( aes(y=median.seg.len, x=Status, group=Status, fill=Status) ) +
  labs(title="Median segment length by patient status")
```

`r pander(sum.patient.data[,c('Patient','Status','years','total.samples','highest.path','n.segments', 'median.seg.len')], justify='left')`

@@

```{r echo=F, message=F, warning=F, fig.height=15, fig.width=10, eval=F}
# Segmented Value Distributions
lapply(patient.data, function(pt) {
  name = pt$info$Patient[1]
  status = pt$info$Status[1]

@@

  if (nrow(sample.normalised.segvals) <= 4) {
    warning(paste(name, 'cannot be run in rcorr, too few observations'))
  }
